An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-07 implementation.tex
fixing bugs

\vspace{1ex}\noindent {\bf Example Properties.}
Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The properties are related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
%
The ASCII version of the logic here uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

>>>>

\vspace{1ex}\noindent {\bf Example properties.}
Figure \ref{fig:properties} shows three properties in the 
input format of the tool, which are not expressible in the original first-order logic of \dejavu{} presented in \cite{HPU}. The properties are related to the examples in Section~\ref{sec:syntax-semantics}, which
are not expressible in (P)FLTL.
%
The ASCII version of the logic here uses 
\idsl{@} for $\ominus$,
\idsl{|} for $\vee$,
\idsl{&} for $\wedge$, and
\idsl{!} for $\neg$.
%
The first property \idsl{telemetry1} is a variant of
formula \ref{eq:wolper-first-order}, illustrating the use of a 
rule to express a first-order version of Wolper's example \cite{Wolper}, that all the states with even indexes of a sequence satisfy a property.
In this case we consider a radio on board a spacecraft, which communicates over different channels (quantified over in the formula), which can be turned on and off with a \idsl{toggle(x)} - they are initially off.
Telemetry can only be sent to ground over a channel \idsl{x} with the \idsl{telem(x)} event when radio channel \idsl{x} is toggled on.

### 2019-02-07 main.tex
fixing bugs

%%%M
\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
One can translate any QLTL property into a second order monadic logic in a standard
way~\cite{Thomas}, and then from
second order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set. A muller automaton 
accepts a run if it 
passes infinitely often in exactly
one of the specified accepting sets. To formulate an equivalent ELTL formula,
the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.

>>>>

%%%M
\subsection*{Details of left to right direction of Theorem~\ref{theo1}.}
One can translate any QLTL property into a second-order monadic logic in a standard
way~\cite{Thomas}, and then from
second-order monadic logic to a deterministic Muller automaton $M = (S, \iota , \Sigma, \Delta, {\cal F})$ (a B\"{u}chi automaton may not always be deterministic), where $S$ is the set of states, $\iota \in S$ is the initial state, $\Sigma$ is the set of inputs, each $\alpha \in \Sigma$ is a minterm (i.e., conjunction of
each variable in the set negated or non-negated) of the variables in $B$, $\Delta : S \times \Sigma \mapsto S$ is the transition function, and ${\cal F} = \{ F_1 , \ldots , F_k \}$ is the accepting set. A muller automaton 
accepts a run if it 
passes infinitely often in exactly
one of the specified accepting sets. To formulate an equivalent ELTL formula,
the auxiliary variables represent the states $S$ of $M$, i.e., $a_s$ represents the state $s \in S$.

### 2019-02-07 rule-ltl.tex
fixing bugs

\vspace{1ex}
\noindent {\bf Related Work.}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds dynamic, i.e.,
state-dependent, quantification over propositions to LTL.

@@

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the 
set of objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
\end{itemize}

@@

The classical definition of linear temporal logic~\cite{MP} has the following syntax:
\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
\, ( \varphi \until \varphi ) \,|  \, \ominus \varphi |
   \, ( \varphi \since \psi ) \,  \]
where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
{\em until}, {\em previous-time} and {\em since}, respectively.

@@

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of different versions of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second order logics where the temporal propositions
correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star operator (or $\omega$).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

@@

\subsubsection{Extending LTL with Dynamic Quantification.}
Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.

@@

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states. It is also called  {\em second order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

@@

\subsubsection{Extending LTL with Rules.}

@@

\noindent
For the other direction, one can translate the QLTL property first into
a second order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton. The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
The details of the
constructions are given in the appendix. \qed

@@

\subsubsection{Syntax}

@@

\subsubsection{Extending FLTL with Dynamic Quantification.}

@@

\subsubsection{Extending FLTL  with Rules.}

@@

\subsubsection{Set Semantics}

@@

\subsubsection{Runtime Verification Algorithm for PFLTL.}

@@

\subsubsection{Runtime Verification Algorithm for EPFLTL}

We extend now the algorithm to capture EPFLTL.
The auxiliary relations $r_j$ extend the model, and we need to keep 
BDDs representing $\current ( r_j )$ and
$\old (r_j)$ for each relation $r_j$. We also need to calculate the subformulas $\varphi_i$ that appear in
a specification, as part of the runtime verification,
as per the above FPLTL algorithm. One subtle point is that the auxiliary relations $r_j$ may be defined in a rule with respect to
a variable $x_j$ (as $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized to any number of variables) נוא $r_j$ can be used
as a subformula with other parameters in other rules or in the statement e.g., as $r_j ( y )$. This can be resolved by a BDD renaming function $\rename ( r_j (x_j ) , y)$.

>>>>

\vspace{1ex}
\noindent {\bf Related work.}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds dynamic, i.e.,
state-dependent, quantification over propositions to LTL.

@@

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second-order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the 
set of objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
\end{itemize}

@@

The classical definition of linear temporal logic~\cite{MP} has the following syntax:
\[ \varphi ::= \true \, | \,  p \, | \, ( \varphi \wedge \varphi ) \, |  \, \neg  \varphi \, |   \, \bigcirc \varphi \, |  
\, ( \varphi  \until  \varphi ) \,|  \, \ominus \varphi |
   \, ( \varphi \ \since \ \psi ) \,  \]
where $p$ is a proposition from a finite set of propositions $P$, and $\bigcirc$, $\until$, $\ominus$, $\since$ stand for {\em next-time}, 
{\em until}, {\em previous-time} and {\em since}, respectively.

@@

%%infinite sequences (finite executions can be extended by repeating the last event indefinitely to form an infinite paths).
The expressive power of different versions of propositional LTL is often compared to regular expressions
over the alphabet $\Sigma = 2^P$ and to  {\em monadic} first and second-order logics where the temporal propositions
correspond to unary predicates (over time). An overview appears in~\cite{Thomas}. Accordingly, we have the following characterizations:
LTL is equivalent to monadic first-order logic\footnote{First-order logic with unary predicates over the naturals, and the relation $<$.},  star-free regular expressions\footnote{Regular expressions but without star operator (or $\omega$).} and counter-free B\"{u}chi automata\footnote{A counter-free language is a regular language for which there is an integer $n$ such that for all words x, y, z and integers $m \geq n$ we have that $xy^m z \in L$ if and only if $xy^n z \in L$.}. In fact, restricting the temporal operators to the {\em future}  operators $\until$ and $\bigcirc$
(and the ones derived from them $\Box$ and $\Diamond$)
leaves the same expressive power and is commonly used.

@@

\subsubsection{Extending LTL with dynamic quantification.}
Adding quantification over propositions, as done in~\cite{Wolper}, allows writing a formula of the form
$\Exists q \, \varphi$, where
$\Exists q$ represents {\em existential} quantification over a proposition $q$ that can appear in $\varphi$.

@@

\begin{itemize}
\item $( \sigma, i ) \models \boldsymbol{\Exists} q \, \varphi$ iff there exists
$\sigma'$ such that $\sigma' |_{\{ q \} } = \sigma$ and 
$( \sigma', i) \models \varphi$.
\end{itemize}
%Since all the auxiliary variables appear in the scope of
%quantifiers, it holds that $\sigma \models \varphi$
%(in $\varphi$, auxiliary variables appear only within
%the scope of quantifiers)
%iff $\sigma |_A \models \varphi$.
%
%\klaus{Are the last two lines above needed?}
%
Universal quantification is also allowed, where $\Forall q\, \varphi =
\neg \Exists q\, \neg \varphi$.
This kind of quantification is considered to be {\em dynamic}, since the quantified variables can have different truth values depending on the states. It is also called  {\em second-order} quantification, since the quantification establishes the {\em set} of states in which a propositional variable has the value $\true$.
Extending LTL with such quantification, the logic QLTL has the same expressive power as regular expressions, full B\"{u}chi
automata, or monadic second-order logic with unary predicates over the naturals, see again~\cite{Thomas}.
In fact, it is sufficient to restrict the quantification to existential quantifiers that prefix the formula to obtain the full expressiveness of QLTL~\cite{Thomas}.

@@

\subsubsection{Extending LTL with rules.}

@@

\noindent
For the other direction, one can translate the QLTL property first into
a second-order monadic logic formula, then to a deterministic Muller automata and then construct an ELTL formula that holds for the accepting executions of this automaton. The rules of this formula encode the automata states, and the statement describes the acceptance condition of the Muller automaton.
The details of the
constructions are given in the appendix. \qed

@@

\subsubsection{Syntax.}

@@

\subsubsection{Extending FLTL with dynamic quantification.}

@@

\subsubsection{Extending FLTL  with rules.}

@@

\subsubsection{Set Semantics.}

@@

\subsubsection{Runtime verification algorithm for PFLTL.}

@@

\subsubsection{Runtime verification algorithm for EPFLTL}

We extend now the algorithm to capture EPFLTL.
The auxiliary relations $r_j$ extend the model, and we need to keep 
BDDs representing $\current ( r_j )$ and
$\old (r_j)$ for each relation $r_j$. We also need to calculate the 
subformulas $\varphi_i$ that appear in
a specification, as part of the runtime verification,
as per the above FPLTL algorithm. One subtle point is that the auxiliary 
relations $r_j$ may be defined in a rule with respect to
a variable $x_j$ (as $r_j (x_j ):= \varphi_j (x_j)$ (this can be generalized 
to any number of variables), but $r_j$ can be used
as a subformula with other parameters in other rules or in the statement e.g., 
as $r_j ( y )$. This can be resolved by a BDD renaming function 
$\rename ( r_j (x_j ) , y)$.
